UD651 Lesson 3 Notes by mrandrewandrade
========================================================

UD651 Udacity's Exploratory Data Analysis Course Notes by Andrew Andrade || mrandrewandrade.com  || March 23rd, 2014 || (CC BY-NC 2.5)


### Lesson 3 Outline:
Lesson 3 teaches us techniques how explore one variable in R.  We take a close look at individual variables in data set, types of variables, distrbution of data, missing data and outliers.

[Installing ggplot and ggplot themes](#P1)     
[Example 1: Setting up gglot ](#E1):  list.files(), read.delim, load ggplot2 and ggthemes library, scale_x_discrete        
[Example 2: Faceting](#E2): facet_wrap and facet_grid in qplot and ggplot     
[Example 3: Limiting and labling axis, adjusting bin width ](#E3): xlim, scale_x_continous     
[Example 4: Faceting friend count into men and women and filtering NA values](#E4): facet_wrap, by     
[Example 5: Playing with color themes and exploring tensure and ages](#E5): color, fill, xlab, ylab     
[Example 6: Transforming data](#E6): log10, sqrt     
[Example 7: Overlaying Frequency Polygons to count total friends and likes](#E7): frequency polygons, overlay plots,  subset       
[Example 8: Box plot](#E8): boxplot, setting coordinatesand limits on boxplot (coord_cartesian)      
[Example 9: Adding new logical variables](#E9): Creating logical variables from what you learnt      


***********

### <a name="P1"></a>Install ggplot and ggplot themes

Run the follow commands in the RStudio terminal:
>install.packages('ggplot2')
>install.packages('ggthemes', dependencies = TRUE) 



### <a name="E1"></a>Example 1: Setting up gglot  
Learn to use: list.files(), read.delim, load ggplot2 and ggthemes library, scale_x_discrete

```{r Pseudo-Facebook User Data}
#Reset the workspace
rm(list = ls(all = TRUE))

#check working directory
getwd()

#set working directory if incorrect
setwd("~/UD651-Course-Notes/Course Notes/lesson3/")

#list files in directory
list.files()

#read.delim defaults to the tab character as the separator between values and the period as the decimal character.
pf <- read.delim('pseudo_facebook.tsv')
names (pf)

#load required libraries
library(ggplot2)
library(ggthemes)
```


```{r fig.width=10, fig.height=10}
#simple histogram of date of births
qplot(x=dob_day, data = pf)+
  scale_x_discrete(breaks=1:31)

```

### <a name="E2"></a>Faceting
Reference: http://www.cookbook-r.com/Graphs/Facets_(ggplot2)/      
Learn to use: facet_wrap and facet_grid in qplot and ggplot


Single Variable Syntax:
facet_wrap(~variable)

Multiple variable Syntax:
facet_grid(vertical~horizontal)

Example using qplot:
```{r fig.width=10, fig.height=10}
#simple histogram of date of births
qplot(x=dob_day, data = pf)+
  scale_x_discrete(breaks=1:31) +
  facet_wrap(~dob_month, ncol=4)

```

Example using ggplot:
```{r fig.width=10, fig.height=10}
#simple histogram of date of births\
ggplot(data = pf, aes(x = dob_day)) +  
  geom_histogram() + 
  scale_x_discrete(breaks = 1:31) + 
  facet_wrap(~dob_month)
  
```


### <a name="E3"></a>Example 3: Limiting and labling axis, adjusting bin width
Reference: http://docs.ggplot2.org/current/scale_continuous.html      
Learn to use:  xlim, scale_x_continous, 

#### Limiting the Axis to 0 to 1000

Using xlim:
```{r fig.width=3, fig.height=3}
#simple histogram of date of births
qplot(x=friend_count, data=pf, xlim = c(0,1000))

```

Using scale_x_continous:
```{r fig.width=3, fig.height=3}
#simple histogram of date of births
qplot(x=friend_count, data=pf) +
  scale_x_continuous(limits =  c(0,1000))

```


Using ggplot:
```{r fig.width=3, fig.height=3}
#simple histogram of date of births
ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram() +
  scale_x_continuous(limits =  c(0,1000))

```

#### Breaks to Set Axis Labels

Using xlim:
```{r fig.width=10, fig.height=10}
qplot(x=friend_count, data=pf, binwidth = 25) +
  scale_x_continuous(limits =  c(0,1000), breaks = seq(0, 1000, 50))

```

Using ggplot:
```{r fig.width=10, fig.height=10}

ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram() + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50))


```


### <a name="E4"></a>Example 4: Faceting friend count into men and women and filtering NA values
Reference: http://www.statmethods.net/input/missingdata.html
Learn to use: facet_wrap, by


Using qplot:

```{r}
qplot(x=friend_count, data=subset (pf, !is.na(gender)), binwidth = 25) +
  scale_x_continuous(limits =  c(0,1000), breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)

```


Using ggplot
```{r}
ggplot(aes(x = friend_count), data = subset (pf, !is.na(gender))) + 
  geom_histogram() + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) + 
  facet_wrap(~gender)

```

We can now look at the total friend count:
```{r}

#using table to compare totals between genders
table(pf$gender)

#using by wrapper to compare summaries of male and female
by(pf$friend_count,pf$gender, summary)

```

### <a name="E5"></a> Example 5: Playing with color themes and exploring tensure and ages
Reference: http://docs.ggplot2.org/0.9.2.1/theme.html
Learn to use: color, fill, xlab, ylab

Using qplot:
```{r}
#color determins the outline of objects
#fill determines the color of area inside object in plot
qplot(x=tenure/365, data = pf, binwidth = 0.25,
      color = I('black'), fill = I('#099DD9')) +
  scale_x_continuous(breaks = seq (1,7,1), limits = c(0,7))+
  xlab('Number of years using Facebook') + 
  ylab('Number of users in sample')
  

```

Using ggplot:
```{r}
ggplot(aes(x = tenure / 365), data = pf) + 
  geom_histogram(color = 'black', fill = '#099DD9') + 
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7)) + 
  xlab('Number of years using Facebook') + 
  ylab('Number of users in sample')
```

User ages:
```{r}
#color determines the outline of objects
#fill determines the color of area inside object in plot
qplot(x=age, data = pf, binwidth = 1,
      color = I('black'), fill = I('#099DD9')) + 
  scale_x_continuous(breaks = seq (0,113,5))
  

```


### <a name="E6"></a> Example 6: Transforming data
Reference:       
Creating multiple plots in one image:      
http://lightonphiri.org/blog/ggplot2-multiple-plots-in-one-graph-using-gridextra       
Add Log or Sqrt Scales to an Axis:     
http://docs.ggplot2.org/current/scale_continuous.html     

Learn to use: log10, sqrt


First install:
>install.packages("gridExtra")

Learn to use:

Transforming friend count data:

```{r}
#Summary of Friend Count
summary (pf$friend_count, data)

#Summary of log of Friend Count
summary (log10(pf$friend_count+1) , data)

#Summary of sqrt of Friend Count
summary (sqrt(pf$friend_count+1) , data)


```

One plot of the transformed data using QPlot:

Using qplot:

```{r}

library(gridExtra) 

p1 <- qplot (x= friend_count, data = pf)
p2 <- qplot (x= log10(friend_count+1), data = pf)
p3 <- qplot (x=  sqrt(friend_count), data = pf)

# arrange plots in grid
grid.arrange(p1, p2, p3, ncol=1)
```

Using ggplot:

```{r}

library(gridExtra) 

p1 <- ggplot (aes(x=friend_count), dat = pf) + geom_histogram()
p2 <-p1 + scale_x_log10()
p3 <- p2 + scale_x_sqrt()

# arrange plots in grid
grid.arrange(p1, p2, p3, ncol=1)
```


### <a name="E7"></a> Example 7: Overlaying Frequency Polygons to count total friends and likes 

Learn to use: frequency polygons, overlay plots,  subset

Using qplot:

```{r}
qplot(x = friend_count, y = ..count../sum(..count..),
      data = subset (pf, !is.na(gender)),
      xlab = 'Friend Count',
      ylab = 'Proportion of Users with that friend count',
      binwidth = 10, geom = 'freqpoly', color = gender) +
  scale_x_continuous (lim = c(0, 1000), breaks = seq(0,1000,200))
      

```


Using ggplot:

```{r}
ggplot(aes(x = friend_count, y = ..count../sum(..count..)), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender)) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 200)) + 
  xlab('Friend Count') + 
  ylab('Percentage of users with that friend count')

```


Visualizing www_likes

Using qplot:

```{r}
qplot (x=www_likes, data = subset (pf, !is.na(gender)),
       geom = 'freqpoly', color = gender) +
  scale_x_continuous() +
  scale_x_log10()


```

Using ggplot:

```{r}
ggplot(aes(x = www_likes), data = subset(pf, !is.na(gender))) + 
  geom_freqpoly(aes(color = gender)) + 
  scale_x_log10()

```


Counting the total likes of men and women

```{r}
#longer solution storing the numbers
males <-subset(pf, gender == 'male')
maleNumberLikes <-sum(males$www_likes)

females <-subset(pf, gender == 'female')
femaleNumberLikes <- sum(females$www_likes)

#shorter solution, just display the numbers
by(pf$www_likes, pf$gender, sum)

```

### <a name="E8"></a> Example 8: Box plot
Reference: http://flowingdata.com/2008/02/15/how-to-read-and-use-a-box-and-whisker-plot/

Learn to use: boxplot, setting coordinatesand limits on boxplot (coord_cartesian)


Using scale_y_continuous or y lim:

```{r}

#using scale y continous
qplot( x = gender, y = friend_count,
       data = subset (pf, !is.na(gender)),
       geom = 'boxplot' ) +
  scale_y_continuous(limits = c(0, 1000))

#using y continous
qplot( x = gender, y = friend_count,
       data = subset (pf, !is.na(gender)),
       geom = 'boxplot', ylim = c(0, 1000))

#using coord_cartesian
qplot( x = gender, y = friend_count,
       data = subset (pf, !is.na(gender)),
       geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 1000))

```

Compare with the summary table:

```{r}

#using coord_cartesian
qplot( x = gender, y = friend_count,
       data = subset (pf, !is.na(gender)),
       geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 250))

by(pf$friend_count, pf$gender, summary)


```

Compare who initiaties more more friendships:

```{r}

by(pf$friendships_initiated, pf$gender, summary)

qplot( x = gender, y =friendships_initiated,
       data = subset (pf, !is.na(gender)),
       geom = 'boxplot') +
  coord_cartesian(ylim = c(0, 1000))
```


### <a name="E9"></a> Example 9: Adding new logical variables

Learn to use: what you have learned to create logical variables


```{r}

pf$mobile_check_in <- NA
pf$mobile_check_in <- ifelse(pf$mobile_likes >0, TRUE, FALSE)
pf$mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)

percentMobileCheckins <- sum(TRUE == pf$mobile_check_in) / length(pf$mobile_check_in) * 100

#display the number of mobile check ins
percentMobileCheckins

```

********
# CONGRATULATIONS
#### You'll now learnt the basics of exploring one variable.  Time to put the knowledge to use!