UD651 Lesson 2 Notes by mrandrewandrade
========================================================

UD651 Udacity's Exploratory Data Analysis Course Notes by Andrew Andrade || mrandrewandrade.com  || March 22nd, 2014 || (CC BY-NC 2.5)

### Lesson 2 Outline:

[Installing R and Rstudio](#P1)      


[Example1 ](#E1): c, nchar   
[Example2 ](#E2): data, names, ?, str, dim, row.names, head, tail, $  
[Example3 ](#E3): getwd, setwd, read.csv, subset, []    
[Example4 ](#E4): plot, summary, boolean operators, add and remove variables, ifelse, rm   
[Example5 ](#E5): factors, table, rearrange table    

[Appendix ](#Appendix):     
Dealing with Untidy Data   
R Markdown    
Free R Resources and Help     

***

### <a name="P1"></a>Install R and RStudio

Download and install the R programming language FIRST at http://cran.rstudio.com. 

After you install R, you can download and install RStudio from http://www.rstudio.com.

RStudio Layout Guide:   
http://dss.princeton.edu/training/RStudio101.pdf

***

### <a name="E1"></a>Example 1: Getting Started
Learn to use: c, nchar


```{r}
#Reset the workspace
rm(list = ls(all = TRUE)) 

# Store names in a vector
names <- c("Chris Saden", "Lauren Castellano",
              "Sarah Spikes","Dean Eckles",
              "Andy Brown", "Moira Burke",
              "Kunal Chawla")

# Display contents of vector
names

# Store 1 to 10 then 1 to 20 in a vector
numbers <- c(1:10)

#display
numbers

#append 11:20
numbers <- c(numbers, 11:20)

#display
numbers
```


#### Display number of characters using nchar:
```{r}
#Reset the workspace
rm(list = ls(all = TRUE))

udacious <- c("Chris Saden", "Lauren Castellano",
              "Sarah Spikes","Dean Eckles",
              "Andy Brown", "Moira Burke",
              "Kunal Chawla", "Andrew Andrade")

# vector that contains the number of characters
# for each of the names in udacious, including my name.
nchar_output = nchar(udacious)
nchar_output

# Check if name in vector contains exactly 11 characters
nchar_output == 11

# Output names with 11 charaters only
udacious[nchar_output  == 11]

```
### <a name="E2"></a>Example 2: Understanding Dataframes

Learn to use: data, names, ?, str, dim, row.names, head, tail, $.

```{r}

#Reset the workspace
rm(list = ls(all = TRUE))

#load mtcars data
data(mtcars)
#display the variable names in mtcars
names(mtcars)
#display help
?mtcars
#display data frame in table form
mtcars

#display structur of data fram
str(mtcars)

#displays the size of the data frame
dim(mtcars)

#help for row names
?row.names

# Change the row names of the cars to numbers.
row.names(mtcars) <- c(1:32)
# Show changes:
mtcars

#reload the data to reset row names
data(mtcars)

#display first 6 entries (default)
head(mtcars)

#display first 10 entries
head(mtcars, 10)

#display last 3 entries
tail(mtcars, 3)

#Access an individual variable (or column) from the data frame
mtcars$mpg
#Find average of that column
mean(mtcars$mpg)

```

#### Note: R uses 1 to start indexing (AND NOT ZERO BASED INDEXING as is true of many other programming languages.)

### <a name="E3"></a>Example 3: Read and Subset Data

Learn to use: getwd, setwd, read.csv, subset, []   
Reference: http://www.statmethods.net/management/subset.html   

```{r}
#Reset the workspace
rm(list = ls(all = TRUE)) 

#Get working directory
getwd()

#Set working directory (change to yours)
setwd('~/UD651-Course-Notes/Course Notes/lesson2/')

#Read in the state info data
statesInfo <-read.csv('stateData.csv')


#One method to store subset the states in region 1 
stateSubset <- subset(statesInfo, 1 == state.region )


#Another method to subset the states in region 1
stateSubsetBraket <- statesInfo[statesInfo$state.region==1, ]


```

### <a name="E4"></a>Example 4: Playing with the cars data set

Learn to use: plot, summary, boolean operators, add and remove variables, ifelse, rm


```{r}
#Reset the workspace
rm(list = ls(all = TRUE)) 

#look at prepacked cars data
?cars

#structure of cars
str(cars)

```

Plot:

```{r fig.width=7, fig.height=6}
plot(cars)
```


```{r}
#show summary of data
summary(mtcars)

#store in efficient the Subset the data to cars that get 23 or more mpg
efficient <- subset (mtcars, mpg >= 23)

#number of cars taht  get more than 23 mpg
dim (efficient)[1]


#subset car(s) that get greater than 30 mpg and have more than 100 raw horsepower.
subset(mtcars, mpg > 30 & hp > 100)

#subset car(s) that get less than 14 mpg or have less more than 390 raw horsepower.
subset(mtcars, mpg < 14 | disp > 390)

# cars that have a 1/4 mile time (qsec) less than or equal to 16.90 seconds to the console.
subset(mtcars, qsec < (16.90+0.25) & qsec > (16.90-0.25) )

# Save the subset of cars that weigh under 2000 pounds (weight is measured in lb/1000) to a variable called lightCars. Print the numbers of cars and the subset to the console.
lightCars <- subset (mtcars, wt < 2)
dim (lightCars)[1]
lightCars

#Create new variables in a data frame (add year)
#assume that all of the models were from 1974.
mtcars$year <- 1974

#remove variable
mtcars <- subset(mtcars, select = -year)
#Notice, we are back to 11 variables in the data frame.

#set years 1973 and 1974 alternating
mtcars$year <- c(1973, 1974)

#remove again
mtcars <-subset (mtcars, select = -year)

#check if the car weight is less than 3
mtcars$wt
cond <- mtcars$wt < 3
cond

#using ifelse

#if true, then set to light
mtcars$weight_class <- ifelse(cond, 'light', 'average')
mtcars$weight_class

#check if the car weight is greater than 3.5
cond <- mtcars$wt > 3.5

#if true then set to heavy
mtcars$weight_class <- ifelse(cond, 'heavy', mtcars$weight_class)
mtcars$weight_class

#remove individual environment variables
rm(cond)
rm(efficient)

```


Quiz: Which car(s) have an mpg (miles per gallon) greater than or equal to 30
OR hp (horsepower) less than 60? Create an R chunk of code to answer the question.

Solution:
```{r}
solution<-subset(mtcars, mpg >= 30 | hp < 60)[0]
```


### <a name="E5"></a>Example 5: Factor Variables
Reference:   
Data type: http://www.statmethods.net/input/datatypes.html   
Reddit Data: http://www.redditblog.com/2011/09/who-in-world-is-reddit-results-are-in.html   

Learn to use: factors, table, rearrange table


```{r}
#Reset the workspace
rm(list = ls(all = TRUE)) 

#Read reddit data
reddit <- read.csv ('reddit.csv')

#summary of data
summary(reddit)

#Check the structure and notice factors
str(reddit)

#table specific factor (employment status)
table(reddit$employment.status)

```


```{r}
#plot the age of users
levels(reddit$age.range)

#use the ggplot library
library(ggplot2)


```


Plot:

```{r fig.width=7, fig.height=6}
qplot(data=reddit, x=age.range)
```

Notice how the x variables on the plot are out of order for Under 18.

Reorder the variables:

Reference: http://statistics.ats.ucla.edu/stat/r/modules/factor_variables.htm

```{r}
reddit$age.range <- ordered(reddit$age.range, levels = c("Under 18", "18-24", "25-34", "35-44", "45-54", "55-64", "65 or Above"))
```

Plot:

```{r fig.width=7, fig.height=6}
qplot(data=reddit, x=age.range)
```

Alternate Solution:

```{r}
reddit$age.range <- factor(reddit$age.range, levels = c("Under 18", "18-24", "25-34", "35-44", "45-54", "55-64", "65 or Above"), ordered = T)
```

Plot:

```{r fig.width=7, fig.height=6}
qplot(data=reddit, x=age.range)
```

# CONGRATULATIONS
#### You'll be exploring data soon with your new knowledge of R.

***
## <a name="Appendix"></a>Appendix

### Dealing with untidy data: Learn data wrangling
https://www.udacity.com/course/ud032   
Tidy Data by Hadley Wickham:   
http://courses.had.co.nz.s3-website-us-east-1.amazonaws.com/12-rice-bdsi/slides/07-tidy-data.pdf   
Tidy Data Presentation by Hadley Whickman (June 2012)   
http://vita.had.co.nz/papers/tidy-data.pdf   

***

### R Markdown    

Code to install knitr for markdown:   

install.packages('knitr', dependencies = T)   
library(knitr)   

Intro Video:   
https://www.youtube.com/watch?v=6A5EpqqDOdk   
Cheat Sheets:      
Help->Markdown Quick Reference   
https://github.com/adam-p/markdown-here/wiki/Markdown-Cheatsheet      
http://warpedvisions.org/projects/markdown-cheat-sheet.md      

Hooks for creating PDF, MS Word etc:    
https://github.com/rstudio/rmarkdown

Code to install
install.packages("devtools")    
devtools::install_github("rstudio/rmarkdown")



### Free R Resources and Help 
http://tryr.codeschool.com/   
https://www.datacamp.com/   
http://www.statmethods.net/  
http://www.cookbook-r.com/  
http://www.r-bloggers.com/  
http://stackoverflow.com/tags/r/info  
http://stackoverflow.com/questions/tagged/r-faq%20  

***